# RStudio exercise 4: Clustering and classification

```{r, include = FALSE}
library(MASS)
library(ggplot2)
library(GGally)
library(corrplot)
```
##Introduction to the data

In this exercise we use Boston data from MASS-library. This dataset contains information collected by the U.S Census Service concerning housing in the area of Boston Mass. Data includes 14 variables and 506 rows.

```{r, echo = FALSE}
dim(Boston)
str(Boston)
```

|variable|description|
|--------|-------------------------------------|
|crim|per capita crime rate by town|
|zn|proportion of residential land zoned for lots over 25,000 sq.ft.|
|indus|proportion of non-retail business acres per town|
|chas|Charles River dummy variable (= 1 if tract bounds river; 0 otherwise)|
|nox|nitrogen oxides concentration (parts per 10 million)|
|rm|average number of rooms per dwelling|
|age|proportion of owner-occupied units built prior to 1940|
|dis|weighted mean of distances to five Boston employment centres|
|rad|index of accessibility to radial highways|
|tax|full-value property-tax rate per $10,000|
|ptratio|pupil-teacher ratio by town|
|black|1000(Bk - 0.63)^2 where Bk is the proportion of blacks by town|
|lstat|lower status of the population (percent)|
|medv|median value of owner-occupied homes in $1000|


###Plotted correlation matrix###

```{r, echo = FALSE}
#calculating the correlation matrix and correlation plot
cor_matrix <- round(cor(Boston),digits = 2)
corrplot(cor_matrix, method="circle", type = "upper", cl.pos = "b", tl.pos = "d", tl.cex = 0.6)
```

Plotted correlation matrix shows that there is some high correlation between variables. Some of them are quite clear as *indus* and *nox*. Industrial adds pollution in the area. *Crim* seems to correlate with *rad* and *tax* but connetion is less clear.  

```{r}
summary(Boston)
```

##Scaled data##
All the variables are numerical so we can use scale()-function to scale whole data set.

```{r, echo = FALSE}
#center and standardize variables
boston_scaled <- scale(Boston)
summary(boston_scaled)
```

```{r}
# class of the boston_scaled object
class(boston_scaled)

# change the object to data frame
boston_scaled <- as.data.frame(boston_scaled)