# RStudio exercise 4: Clustering and classification

```{r, include = FALSE}
#install.packages("plotly")
library(MASS)
library(ggplot2)
library(GGally)
library(corrplot)
library(dplyr)
library(plotly)
```
##Introduction to the data

In this exercise we use Boston data from MASS-library. This dataset contains information collected by the U.S Census Service concerning housing in the area of Boston Mass. Data includes 14 variables and 506 rows.

```{r, echo = FALSE}
dim(Boston)
str(Boston)
```

|variable|description|
|--------|-------------------------------------|
|crim|per capita crime rate by town|
|zn|proportion of residential land zoned for lots over 25,000 sq.ft.|
|indus|proportion of non-retail business acres per town|
|chas|Charles River dummy variable (= 1 if tract bounds river; 0 otherwise)|
|nox|nitrogen oxides concentration (parts per 10 million)|
|rm|average number of rooms per dwelling|
|age|proportion of owner-occupied units built prior to 1940|
|dis|weighted mean of distances to five Boston employment centres|
|rad|index of accessibility to radial highways|
|tax|full-value property-tax rate per $10,000|
|ptratio|pupil-teacher ratio by town|
|black|1000(Bk - 0.63)^2 where Bk is the proportion of blacks by town|
|lstat|lower status of the population (percent)|
|medv|median value of owner-occupied homes in $1000|

##Graphical overview of the data##

###Plot matrix of the data###

```{r, echo = FALSE}
pairs(Boston, col = 4)
```

There are some very intresting distributions fo variables in the plot matrix. Variable *rad* has high and low values so the plot shows that the values are consenrated either side of the plot. VAriable *   

###Plotted correlation matrix###

```{r, echo = FALSE}
#calculating the correlation matrix and correlation plot
cor_matrix <- round(cor(Boston),digits = 2)

corrplot(cor_matrix, method="circle", type = "upper", cl.pos = "b", tl.pos = "d", tl.cex = 0.6)
```

Plotted correlation matrix shows that there is some high correlation between variables:

* Correlation is quite clear between industrial areas (*indus*) and nitrogen oxides (*nox*). Industry adds pollution in the area. Industry seems to correlate also with variablrs like *age*, *dis*, *ras* and *tax*.

* Nitrogen oxides (*nox*) correlations are very similar with industry (*indus*)

* Crime rate (*crim*) seems to correlate with good accessibilitty to radial highways (*rad*) and value property (*tax*).

* Old houses (*age*) and employment centers have also something common 



```{r}
summary(Boston)
```

##Scaled data##
All the variables are numerical so we can use scale()-function to scale whole data set.

```{r, echo = FALSE}
#center and standardize variables
boston_scaled <- scale(Boston)
summary(boston_scaled)
class(boston_scaled)
boston_scaled <- as.data.frame(boston_scaled)
```

Scaling the data makes variables look as if they are in the same range. Variables like *black* and *tax* were before scaling hundred fold compared to some other variables. 

##Creating a new categorical variable crime##

Variable *crim* is the base of the new categorical variable *crime*. 

|categories|quantile points|
|-------|-------|
|low|0%-25%|
|med_low|25%-50%|
|med_high|50%-75%|
|high|75%-100%|

```{r, echo = FALSE}
#save the scaled crim as scaled_crim
scaled_crim <- boston_scaled$crim
```

Quantile points of the variable *crim*

```{r, echo = FALSE}
#create a quantile vector of crim and print it
bins <- quantile(scaled_crim)
bins

#create a categorical variable 'crime'
crime <- cut(scaled_crim, breaks = bins, include.lowest = TRUE, label = c("low", "med_low", "med_high", "high"))

#look at the table of the new factor crime, do not change the actual variable "crime"
crime_tab <-table(crime)
crime_tab

#remove original crim from the dataset
boston_scaled <- dplyr::select(boston_scaled, -crim)

# add the new categorical value to scaled data
boston_scaled <- data.frame(boston_scaled, crime)
summary(boston_scaled)
```

##Train and test sets##

Training set contains 80% of the data. 20% is in the test set.

```{r, echo = FALSE}
# number of rows in the Boston dataset 
n <- nrow(boston_scaled)

# choose randomly 80% of the rows
ind <- sample(n,  size = n * 0.8)
ind
# create train set
train <- boston_scaled[ind,]

# create test set 
test <- boston_scaled[-ind,]
```

##Fitting the Linear Discriminant Analysis##

First the linear discriminant analysis (LDA) is fitted to the train set. The new categorical variable *crime* is the target variable and all the other variables of the dataset are predictor variables.

After fitting we draw the LDA biplot with arrows.

```{r, echo = FALSE}
#LDA = linear discriminant analysis
lda.fit <- lda(crime ~. , data = train)

#print the lda.fit object
lda.fit

#the function for lda biplot arrows
lda.arrows <- function(x, myscale = 1, arrow_heads = 0.1, color = "red", tex = 0.75, choices = c(1,2)){
  heads <- coef(x)
  arrows(x0 = 0, y0 = 0, 
         x1 = myscale * heads[,choices[1]], 
         y1 = myscale * heads[,choices[2]], col=color, length = arrow_heads)
  text(myscale * heads[,choices], labels = row.names(heads), 
       cex = tex, col=color, pos=3)
}

#target classes as numeric
classes <- as.numeric(train$crime)
classes

#plot the lda results
plot(lda.fit, dimen = 2, col = classes, pch = classes)
lda.arrows(lda.fit, myscale = 3)

```

##Predicting the classes##

```{r, echo = FALSE}
#save the correct classes from test data
correct_classes <- test$crime

#remove the crime variable from test data
test <- dplyr::select(test, -crime)

#predict classes with test data
lda.pred <- predict(lda.fit, newdata = test)

#cross tabulate the results
table(correct = correct_classes, predicted = lda.pred$class)
```

Prediction were quite good. There was some errors in the middle of the range but classes low and especially high were good. Only one correct representative of high class was predicted to med_low class. 

##K-means algorithm##

I'm going to calculate what is the optimal number of clusters for Boston data. First I reload and scale the data. Variables need to be scaled to get comparable distances between observation. 
```{r, echo = FALSE}
#Loading and scaling Boston data
scaled_Boston <- scale(Boston)
summary(scaled_Boston)
```

Next I calculate the distances between observations and determinen the number of clusters.
```{r, echo = FALSE}
#calculating the euclidean distance matrix between the observation
dist_eu <- dist(scaled_Boston)

#determining the max number of clusters
cluster_max <- 15

#calculate the total within sum of squares
#K-means might produce different results every time, because it randomly 
#assigns the initial cluster centers. The function set.seed() can be used to 
#deal with that.
set.seed(123)
twcss <- sapply(1:cluster_max, function(k){kmeans(dist_eu, k)$tot.withinss})

# visualize the results
plot(1:cluster_max, twcss, type='b')
```

One way to determine the number of clusters is to look how the total of within cluster sum of squares (WCSS) behaves when the number of clusters changes. WCSS was calculated from 1 to 15 clusters. The optimal number of clusters is when the total WCSS drops radically. It seems that in this case optimal number of clusters is two. However we are here to learn so I decided to analyse model with four clusters. 

After determining the number of clusters I run the K-means alcorithm again.
```{r, echo = FALSE}
#k-means clustering
km <-kmeans(dist_eu, centers = 4)

# plot the Boston dataset with clusters
pairs(Boston, col = km$cluster)
```

It seems that when the data is divided to four clusters there is some clear differences in distriputions of several variables. *Crim*, *zn*, *indus* and *blacks* are variables were one can distinguish clear patterns between clusters. Some variables (*rad* & *tax*) show that sometimes 1 or 2 clusters make a clear distripution but observation of other two clusters are ambigious and there is no clear pattern to be regognised.

##BONUS: LDA using clusters as target classes##

After loading the Boston dataset I scale it to get comparable distances. 
```{r, echo = FALSE}
#Loading and scaling Boston data
scaled_Boston <- scale(Boston)
scaled_Boston <- as.data.frame(scaled_Boston)
#colnames(scaled_Boston)

#calculating the euclidean distance matrix between the observation
dist_eu <- dist(scaled_Boston)

#k-means clustering
set.seed(123)
km <-kmeans(dist_eu, centers = 4)
cm <- as.data.frame(km$cluster)

#adding the clusters to the scaled dataset
scaled_Boston <- data.frame(scaled_Boston, clust = cm)
colnames(scaled_Boston)[15] <- "clust"

summary(scaled_Boston)
```

Original Boston dataset is now scaled and the result of K-means clustering is saved to the variable *clust*

###LDA with the clusters###

Next the LDA is performed and the biplot with arrows is created
```{r, echo = FALSE}
#LDA = linear discriminant analysis
lda.fit.km <- lda(clust ~. , data = scaled_Boston)

#print the lda.fit object
lda.fit.km

#the function for lda biplot arrows
lda.arrows <- function(x, myscale = 1, arrow_heads = 0.1, color = "red", tex = 0.75, choices = c(1,2)){
  heads <- coef(x)
  arrows(x0 = 0, y0 = 0, 
         x1 = myscale * heads[,choices[1]], 
         y1 = myscale * heads[,choices[2]], col=color, length = arrow_heads)
  text(myscale * heads[,choices], labels = row.names(heads), 
       cex = tex, col=color, pos=3)
}

#target classes as numeric
classes <- as.numeric(scaled_Boston$clust)
#classes

#plot the lda results
plot(lda.fit.km, dimen = 2, col = classes, pch = classes)
lda.arrows(lda.fit.km, myscale = 3)
```

Biplot shows that variables *indus*, *zn* and *medv* are the most influencial separators for the clusters.

##Super-bonus##

###3D plot where observations color is the crime classes of the train set###
```{r, echo = FALSE}
model_predictors <- dplyr::select(train, -crime)

#check the dimensions
#dim(model_predictors)
#dim(lda.fit$scaling)

#matrix multiplication
matrix_product <- as.matrix(model_predictors) %*% lda.fit$scaling
matrix_product <- as.data.frame(matrix_product)

#3d plot
plot_ly(x = matrix_product$LD1, y = matrix_product$LD2, z = matrix_product$LD3, type= 'scatter3d', mode='markers', color = train$crime)
```

###3D plot where observations color is based on the K-means clusters.###
```{r, echo = FALSE}
model_predictors <- dplyr::select(scaled_Boston, -clust)

#check the dimensions
#dim(model_predictors)
#dim(lda.fit.km$scaling)

#matrix multiplication
matrix_product <- as.matrix(model_predictors) %*% lda.fit.km$scaling
matrix_product <- as.data.frame(matrix_product)

#3D plot
plot_ly(x = matrix_product$LD1, y = matrix_product$LD2, z = matrix_product$LD3, type= 'scatter3d', mode='markers', color = scaled_Boston$clust)
```


Colors of the both plots is based to four classes. It seems that K-means plot shows the different clusters more clearly than the plot that is based on the crime classification. 
