# RStudio exercise 3: Logistic regression
```{r setup,  include = FALSE}
library(ggplot2)
library(GGally)
```
##Introduction to the data
Using Data Mining To Predict Secondary School Student Alcohol Consumption. 
Fabio Pagnotta, Hossain Mohammad Amran 
Department of Computer Science,University of Camerino

https://archive.ics.uci.edu/ml/datasets/STUDENT+ALCOHOL+CONSUMPTION

Data is rolled into one from Math course and Portuguese language course datasets. After the data wrangling exercise the new data set is found from the data folder.
```{r}
student_alc <- read.table("data/student_alc.txt", header = TRUE)
dim(student_alc)
```
The student data includes 35 variables and 382 rows.
```{r}
colnames(student_alc)
```
* The variables not used for joining the two data have been combined by averaging (including the grade variables)
* 'alc_use' is the average of 'Dalc' and 'Walc'
* 'high_use' is TRUE if 'alc_use' is higher than 2 and FALSE otherwise

##High and low alcohol consumption and other variables

In this analysis I'm going to study the relationship of high/low alcohol consumption and the following variables:

* age
* studytime
* freetime
* absence

First these relationships are observed from tables and graphics

```{r, echo = FALSE}
# initialize a plot of high_use and sex
g1 <- ggplot(student_alc, aes(x = high_use, y = age, col = sex))

# define the plot as a boxplot and draw it
g1 + geom_boxplot() + ylab("Age by alcohol consumption and sex")
```

a

