# RStudio exercise 5: Dimensionality reduction techniques

```{r}
library(MASS)
library(ggplot2)
library(GGally)
library(corrplot)
library(dplyr)
library(plotly)
```

##Introduction to the data

The data of this exercise originates from the United Nations Development Programme. Human development index (HDI) was created to emphasize that people and their capabilities should be the ultimate criteria for assessing the development of a country, not economic growth alone.

Original data can be found from: http://hdr.undp.org/en/content/human-development-index-hdi

Modified dataset includes 9 variables and 155 observations as follows:

|variable|description|
|----------|-----------|
|country|Country name (name of the row)|
|gni_capita|Gross national income per capita|
|life_exp|Life expectancy at birth|
|edu_years|Expected years of schooling|
|mortality|Maternal mortality ratio|
|young_mom|Adolescent birth rate|
|women_parlament|Percentange of female representatives in parliament|
|edu_ratio|Ratio between females and males at least secondary education| 
|labour_ratio|Ratio between females and males in the labour force|


###Summary of the data###
```{r, echo = FALSE}
human <- read.table("data/human.txt", header = TRUE)
str(human)
```

All the variables are either numerical or integers. Two of the variables are ratios (*edu_ratio*, *labour_ratio*).

###Summary, distributions of the variables and the relationships between them###
```{r, echo = FALSE}
summary(human)
```

There is greate variation between numeric variables. Variable like Gross national income per capita (*gni_capita*) differs between 581$ (min) -123,124$ (max) and at the same time there is two variables (*edu_ratio*, *labour_ratio*) that are ratios and differs both sides of number 1. Life expectancy (*life_exp*) and  expected years of schooling (*edu_years*) are in years and the max observation is under 100 years. The percentange of female representatives in parliament (*women_parlament*) is the only variable that is represented in percentange.


```{r, echo = FALSE}
ggpairs(human, mapping = aes(alpha = 0.1), lower = list(combo = wrap("facethist", bins = 20)))
``` 


