# RStudio exercise 3: Logistic regression
```{r,  include = FALSE}
library(ggplot2)
library(GGally)
library(dplyr)
```
##Introduction to the data
Using Data Mining To Predict Secondary School Student Alcohol Consumption. 
Fabio Pagnotta, Hossain Mohammad Amran 
Department of Computer Science,University of Camerino

https://archive.ics.uci.edu/ml/datasets/STUDENT+ALCOHOL+CONSUMPTION

Data is rolled into one from Math course and Portuguese language course datasets. After the data wrangling exercise the new data set is found from the data folder.
```{r}
alc <- read.table("data/student_alc.txt", header = TRUE)
dim(alc)
```
The student data includes 35 variables and 382 rows.

* The variables not used for joining the two data have been combined by averaging (including the grade variables)
* 'alc_use' is the average of 'Dalc' and 'Walc'
* 'high_use' is TRUE if 'alc_use' is higher than 2 and FALSE otherwise

##High and low alcohol consumption and other variables


In this analysis I'm going to study the relationship of high/low alcohol consumption between sex and the following variables:

* age
* studytime
* freetime
* absence

First these relationships are observed from tables and graphics. Hypothesis are as follows:

**Age**

The age of high consumption of alcohol may differ between sex. The development of charcter differs between young people and this may affect on habbits of alcohol consumption.
H0:Age don't affect on alcohol consumption
H1:There is difference in between level of alcohol consumption and age  

```{r}
#a jitterplot of high_use and sex
g1 <- ggplot(alc, aes(x = high_use, y = age, col = sex))
g1 + geom_jitter() + ggtitle("Age by alcohol consumption and sex")
```

It seems that there is randomnes of sex and age in both alcohol consumption groups.

```{r}
#a boxplot of high_use, sex and age
g1 <- ggplot(alc, aes(x = high_use, y = age, col = sex))
g1 + geom_boxplot() + ggtitle("Age by alcohol consumption and sex")
```

Means however show that young male students have lower mean of age in low consumption group and females in high consumption group.



**Studytime**

High alcohol consumption may be related to time spend in studies because one can't do both at the same time at least not successfully.
H0:Group of high alcohol consumption affects on wekkly study time (numeric: 1 - <2 hours, 2 - 2 to 5 hours, 3 - 5 to 10 hours, or 4 - >10 hours) 

```{r}
#barplot about studytime
ggplot(alc, aes(studytime, fill=high_use)) + geom_bar(position="dodge") +
  ggtitle("Barplot about studytime grouped by high_use")+ xlab("Study time") + ylab("Frequency") +
  theme(plot.title = element_text(hjust = 0.5,size=16,face='bold'),
        axis.title= element_text(hjust = 0.5,size=14))
```


