# RStudio Exercise 2: Analysis

<!--/*Describe the work you have done this week and summarize your learning.*

- Describe your work and results clearly. 
- Assume the reader has an introductory course level understanding of writing and reading R code as well as statistical methods
- Assume the reader has no previous knowledge of your data or the more advanced methods you are using  /-->

##Introduction to the data
After the data wrangling exercise the new data set is found from the data folder. The set is based on data that was collected from course Introduction to Social Statistics, fall 2014 - in Finnish. The survey was conducted  3.12.2014 - 10.1.2015 by Kimmo Vehkalahti.
```{r}
student2014 <- read.table("data/learning2014.txt", header = TRUE)
dim(student2014)
```
The student data includes 7 variables and 166 rows.
```{r}
str(student2014)
```
Variables deep, strat and surf are combination variables from the original survey data.

```{r}
summary(student2014)
```
From 166 survey respondents 56 was men and 110 was females. The mean of age was 25,5 years. Oldes respondet was 55 and youngest 17 years old.

##Graphical output
```{r setup,  include = FALSE}
library(ggplot2)
library(GGally)
```

```{r, echo = FALSE}
ggpairs(student2014, mapping = aes(col = gender, alpha = 0.1), lower = list(combo = wrap("facethist", bins = 20)))
```  

Variables differ between genders. Distributions are different in age, attitude and surf (surfface). Three highest correlation between variables are:

* points-attitude
* surf-deep
* surf-attitude


```{r, echo = FALSE}
qplot(attitude, points, data = student2014, main = "Scatter plot between attitude and points") + geom_smooth(method = "lm")
```

##Explanatory variables
###Three variables
```{r, echo = FALSE}
my_model<- lm(points ~ attitude + stra + surf, data = student2014)
summary(my_model)
```

* In this linear regression model points are the target variable and attitude, strategy (stra) and surfface (surf) are explanatory variables.

* Residuals of the model are between ~ -17.2 and ~10.9 when median is 0.52. I assume that errors are normally distributed but distribution needs confirmation.

* Attitude is the only variable that has a very good significance in this model. p-value of stra and surf is too high to be even slightly significance.  

* Variables estimated coefficient is ~0.34 and it's standard error is clearly smaller (~0.057). Other explanatory variables have not significance in this model.
 
* Residual standard error is high in relation to a first and third quantiles of residuals. 

* This linear regression model of three explanatory variables explains ~19.3% of the points.

###One variable
```{r, echo = FALSE}
my_model2<- lm(points ~ attitude, data = student2014)
summary(my_model2)
``` 
 
* Estimated coefficient of explanatory variable attitude is ~0.35. This means that when attitude rises one point target variable (points) grows 0.35 times.

* Attitude p-value (0.00000000412) shows that variable is very significant in this linear regression model

* This model explains 18.6% of the points (target variable)

##Graphical model validation
```{r, echo = FALSE}
plot(my_model2, which = c(1:2, 5) )
```

**Residuals vs Fitted**
This plot shows that there is no pattern between residuals. There is a constant variance among errors. One can confirm that the assumption of constant variance of errors is valid.  

**Normal Q_Q**
Normal QQ -plot shows that the errors of the model are normally distributed.

**Residual vs Leverage**
Last plot of the graphical model validation shows that the impact of the singel observation is moderate. Model includes some outliers but the leverage of singel observation don't compromise the validation of the model.